Water desalination using polyelectrolyte hydrogel. Gibbs ensemble modelling

### 2022-06-30 gibbs.tex

\newcommand{\Ka}{K_{\mathrm{A}}}
\newcommand{\pKa}{\mathrm{p}\Ka}
\newcommand{\pK}{\mathrm{p}K}
\newcommand{\pH}{\mathrm{pH}}
\newcommand{\mol}{\mathrm{mol}}
\newcommand{\molperl}{\mathrm{mol/l}}
\newcommand{\kg}{\mathrm{kg}}
\newcommand{\res}{^{\mathrm{res}}}
\newcommand{\pHres}{\pH\res}
\newcommand{\pHgel}{\pH\gel}
\newcommand{\cs}{c_{\mathrm{s}}}
%\newcommand{\cs}{c_{\mathrm{salt}}}
\newcommand{\csres}{\cs\res}
\newcommand{\Vgel}{V\gel}
\newcommand{\Vgeleq}{\widetilde{V}\gel}
\newcommand{\Pgel}{P\gel}
\newcommand{\Pres}{P\res}
\newcommand{\Vout}{V\out}
\newcommand{\Vbox}{V\tot}
\newcommand{\PE}{polyelectrolyte{}}

@@

\begin{document}
\begin{abstract}
Recently polyelectrolyte hydrogels have been proposed as draw agents
for reverse osmosis desalination techniques. Indeed, polyelectrolyte
hydrogels have the ability to absorb a big amount of water across
forward osmosis membrane as a result of their swelling pressure. The
insoluble cross-linked network of the gel enables dewatering under
the influence of a stimuli (thermal and/or mechanical). On the other
hand, the network structure of a polymer hydrogel from a thermodynamic
perspective is already an osmotic membrane. So hydrogel microparticles
may allow to completely avoid the osmotic membranes in forward osmosis
and use microfiltration instead. By this article, we present our recent
theoretical study of the use of polyelectrolyte hydrogel for water
desalination. We modeled the thermodynamic equilibrium of coexistence
of the gel and the aqueous salt solution in the so-called closed ensemble,
in which the total amount of ions is assumed to be constant. We modeled
the compression of the gel and the associated with that release of
the solution. We have shown that the squeezed out solution has a little
lower salinity than that the gel was equilibrated with. Also, we performed
a set of simulations modeling the process of continuous decrease of
water salinity up to freshwater concentrations.

\end{abstract}

@@

The first approach~--- distillation. It uses thermal means to effect
a phase change of the water to vapor (or solid), then to physically
separate the vapor phase from the remaining salt solution, and then
recover the thermal energy for reuse as the separated vapor reverts
to liquid form. Distillation processes were the first desalination
techniques conducted on a large commercial scale and still account
for a large portion of the modern world’s desalination capacity.

The second approach is to physically separate the brine components,
generally with a membrane, as they move in response to an externally
applied gradient. In the context of our study, we will mention the
reverse osmosis process (RO) \textasciitilde -{}-{}- the major processes
of all the modern desalination industry, and the new emerging membrane
technology -{}-{}- forward osmosis (FO) \cite{Akther2015}. In RO,
water passes through a membrane that is impermeable to the solute
in response to a chemical potential gradient achieved through pressurization.
In FO, a concentrated draw solution, with a lower water chemical potential,
and a feed solution, with a higher water chemical potential, are separated
by a membrane that rejects the salt but allows the passage of water.
The water permeation from feed solution to draw solution is a spontaneous
process driven by the chemical potential gradient.

Distillation is easy and cheap technology but it is characterized
by relatively high energy costs due to the dissipation of thermal
energy. On the contrary, RO uses expensive osmotic membranes, which
need to be replaced regularly because of scaling and fouling. Moreover,
RO requires very high operating pressures, ranging from 20 to 200
bar, to let the water pass through the membrane. But in terms of energy
losses, it works close to the thermodynamic limit. Thus, theoretically,
per one transferred from freshwater to salty water ion pair, it consumes
only the energy of difference of those ions chemical potentials.

The absence of large hydraulic pressure in FO process in comparison
to RO allows to reduce the energy consumption in pumping, reduce membrane
scaling and fouling, therefore significantly increase the lifetime
of the membranes. In FO, the draw solutes (agents) dispersed and/or
dissolved in water form homogeneous draw solutions. The correct choice
of the draw agents is a task of paramount importance. As an osmotically
driven process, the draw solute is expected to significantly reduce
the water chemical potential, and consequently generate a high osmotic
pressure. On the other hand, the diluted draw solution is also expected
to be easily separated to fresh water and to concentrated draw solution
for reuse \cite{Cai2016}.

\paragraph{Hydrogels for desalination.}

Hydrogels are three-dimensional networks of polymer chains that are
crosslinked by either physical or chemical bonds. They are able to
entrap large volumes of water attracted by the high concentration
of hydrophilic groups. Hydrogels with ionic groups on the comonomer
unit are able to reject salt ions from the solution, they absorb a
solution of lower salinity than that they are equilibrated with. When
a dehydrated or deswollen hydrogel uptakes water, its polymer chains
extend, creating a swelling pressure. For example, as reported in
\cite{Wack_2009} weakly crosslinked poly(acrylic acid)/poly(sodium
acrylate) copolymers with polymer volume fractions between 0.03 and
0.30 exhibit a swelling pressure ranging from 0.20--4.23 MPa.

An important, advantageous aspect of polymer hydrogels is that they
can undergo reversible volume change, \ie solution --- gel phase
transitions in response to environmental stimuli. This aspect causes
hydrogels to be labeled as ‘smart’ hydrogels. Many physical and chemical
stimuli have been applied to induce various responses of such smart
hydrogels, in particular, to change them from hydrophilic to hydrophobic,
thereby releasing water. These stimuli include: temperature, solvent
composition, light, mechanical pressure, sound, electric and/or magnetic
field, whilst the chemical or biochemical stimuli include pH, ions,
and specific molecular recognition events \cite{Tanaka_1982,Serizawa_2001,Lietor_Santos_2009,Qiu_2001}.

The smart properties of hydrogels were utilized for desalination in
a recent study by Li \etal \cite{Li2011}, where hydrogels have been
considered as draw agents for FO. The main point of the cited article
is that hydrogels, due to their swelling and osmotic pressure, were
able to absorb water across the FO membrane. The insoluble cross-linked
network of the gel enables dewatering under the influence of stimuli
(thermal and/or mechanical). For instance, Li \etal proposed the
use of hydrogel made of thermoresponsive polyelectrolyte~--- copolymer
of poly-N-isopropyl acrylamide (p-NIPAAm) and polyacrylic acid (p-AA).
Depending on the temperature the network of the gel may appear to
be either hydrophilic or hydrophobic. In the hydrophilic state, the
gel is used as a draw agent accumulating water inside, then in the
hydrophobic state, the water releases out.

Moreover the hydrogel can be as an osmotic membrane from a thermodynamics
perspective \cite{Wang_2014}. Such a view on hydrogels was employed
in a series of works by the group of prof. Wilhelm (see for example
\cite{Arens_2017,Fengler_2020}). The authors proposed to get rid of
the osmotic membrane and simply use only a micro filtration membrane
to compress the hydrogel and separate it from the solution. In their
method first, the deswollen hydrogel was equilibrated with saline
water feed. During equilibration, the gel was swelling absorbing water.
Then it was taken out from the feed solution and mechanically compressed
by means of a microfiltration membrane. The squeezed out brine was
obtained to have lower salinity than the feed water.

A similar approach was used by Ali \etal \cite{Ali2015}. In their
study the thermosensitive gel was used (based on copolymers p-NIPAAm
and p-AA), so instead of physical compression, the dewatering was
done by an external heating stimulus (sunlight). At night the gel
was equilibrating with feed water and at daytime, under sunlight,
the gel was shrinking and releasing the collected inside water.

@@

The maximum entropy principle requires the density of charges to be
more-less the same throughout the system medium. Since the polyelectrolyte
gel has its own charges and its own neutralizing counterions, the
density of mobile ions (which are able to freely enter and leave the
gel) in gel appears to be smaller than their density outside the gel.
Therefore the internal solution of the gel has lower density of mobile
ions than the solution outside. In that sense, the gel acts as an
osmotic membrane separating solutions \cite{Rud2020}. The driving
force of the separation is the Donnan potential originated from the
charges of the hydrogel network. The density of mobile ions difference
between ions concentrations in the internal and external solutions
is defined by Donnan law \cite{Rud2018}
\[
\frac{c_{-}^{in}}{\cs}=\frac{\cs}{c_{+}^{in}}=\sqrt{1+\left(\frac{\alpha\cp}{2\cs}\right)^{2}}\pm\frac{\alpha\cp}{2\cs}
\]
where $c_{-}^{in}$ and $c_{+}^{in}$ are the internal concentration
of anions and cations, $\cs$ is the salinity of external solution
$\cp$ and $\alpha$ are the gel molar density and the ionization
degree. The $\pm$ sign in the formula depends on whether the gel
is made of polyanion or polycation.

The internal solution can be extracted by means of compression and/or
other stimulus providing that the charge of the gel remains constant.
In the case of gel made of weak (pH-sensitive) polyelectrolyte, the
compression initiates discharging of the gel, therefore the neutralizing
counterions may leave out the gel diminishing the desalination effect
and affecting the pH of the extracted solution. Nevertheless, the
release of counterions may be employed for desalination as well and
the way how to do it is discussed in\cite{Rud2018}. 
\begin{enumerate}
\item However, one can argue that sole Donnan effect is insufficient to
achieve a high salt rejection \cite{Cai2016} and the salinity of
the water squeezed from hydrogel under a very high hydraulic pressure
(up to 100 bar \cite{Fengler_2020}) turns to be not much different
from the initial.
\item Indeed, the use of high hydraulic pressure diminishes all the advantages
of this method over the reverse osmosis. At the same time the reversibility
of hydrogel swelling after such strong compression remains questionable. 
\item Nevertheless by the current study we would like to model the compression
of the gel limiting ourselves within low compression rates, less than
5 bar. We will show how the compression of the gel affects the surrounding
salinity and model the whole desalination process as a cascade of
$c$. 
\end{enumerate}
In the current study we would like to concentrate on the use of gel
made of strong polyelectrolyte for water desalination. Namely, to
investigate how the compression of the gel affects the salinity of
the surrounding the gel solution and to propose the way of water desalination
employing the hydrogel swelling/deswelling set of processes.

The target of our study is to model the thermodynamic equilibrium
between the two boxes and to show how the compression of the gel affects
the ion density in the outside solution.

@@

\begin{figure*}[t]
\subfloat[Partial pressure of the gel versus gel molar volume \label{fig: PV}]
{\includegraphics[width=0.48\textwidth]{figures/fig_PV_pub}}
\hspace{0.02\textwidth}
\subfloat[Supernatant salinity versus gel molar volume\label{fig: CV}]
{\includegraphics[width=0.48\textwidth]{figures/fig_CV_pub}}
\caption{
The compression of the gel in open system (solid lines) and in closed system (dotted lines). 
Each solid curve corresponds to different salinity of the reservoir $\cs$ (see legend). Shadowed area limit the states with applied pressure below zero and above 5 bar.\label{fig: PV and CV}}
\todoi{replace $V$ by $\Vgel$ in Figures}
\end{figure*}

@@

\begin{figure*}[t]
\subfloat[Pressure vs gel density \label{fig: NV}]
{\includegraphics[width=0.49\textwidth]{figures/fig_NV_pub}}
\subfloat[Salinity vs gel density\label{fig: CN}]
{\includegraphics[width=0.495\textwidth]{figures/fig_NVbox_mu_pub}}
\caption{The compression of the gel in open system (solid lines) and in closed
system (dotted lines). Each solid curve corresponds to different salinity
of the reservoir $\cs$ (see legend). Shadowed area limit the states
with applied pressure below zero and above 5 bar.\label{fig: NV and CN}}
%\todoi{Replace y-axis labels by $\ncl$}
\end{figure*}

@@

\caption{ORIGINAL TABLE (to be removed)
}
\end{table}
\end{document}

>>>>

\newcommand{\Ka}{K_{\mathrm{A}}}
\newcommand{\pKa}{\mathrm{p}\Ka}
\newcommand{\pK}{\mathrm{p}K}
\newcommand{\pH}{\mathrm{pH}}
\newcommand{\mol}{\mathrm{mol}}
\newcommand{\molperl}{\mathrm{mol/l}}
\newcommand{\kg}{\mathrm{kg}}
\newcommand{\res}{^{\mathrm{res}}}
\newcommand{\pHres}{\pH\res}
\newcommand{\pHgel}{\pH\gel}
\newcommand{\cs}{c_{\mathrm{s}}}
%\newcommand{\cs}{c_{\mathrm{salt}}}
\newcommand{\csres}{\cs\res}
%\newcommand{\Vgel}{V\gel}
\newcommand{\Vgel}{v_\mathrm{gel}}
\newcommand{\Vgeleq}{\widetilde{\Vgel}}
\newcommand{\Pgel}{P\gel}
\newcommand{\Pres}{P\res}
\newcommand{\Vout}{V\out}
\newcommand{\Vbox}{V\tot}
\newcommand{\PE}{polyelectrolyte{}}

@@

\begin{document}
\begin{abstract}
Recently polyelectrolyte hydrogels have been proposed as draw agents for reverse osmosis desalination techniques. 
Indeed, polyelectrolyte hydrogels have the ability to absorb a big amount of water across forward osmosis membrane as a result of their swelling pressure. 
The insoluble cross-linked network of the gel enables dewatering under the influence of a stimulus (thermal and/or mechanical) \todo{remove this sentence?}.
On the other hand, the network structure of a polymer hydrogel, from a thermodynamic
perspective, is already an osmotic membrane. So hydrogel microparticles
may allow to completely avoid the osmotic membranes in forward osmosis
and use microfiltration instead.

In this article, we present our recent theoretical study of the use of polyelectrolyte hydrogel for water desalination. 
We modeled the thermodynamic equilibrium of coexisting gel and aqueous salt solution phases in the so-called `closed' ensemble, in which the total amount of ions is assumed to be constant. 
We modeled the compression of the gel associated with the release of solution and have shown that the squeezed-out solution has a little lower salinity than the one the gel was equilibrated with. 
Finally, we performed a set of simulations modeling the process of continuous decrease of water salinity up to freshwater concentrations.
\end{abstract}

@@

The first approach is distillation, which uses heat to cause a phase change of the water to vapor. The vapor phase is separated from the brine and condenses to liquid fresh water. 
The released condensation energy is directed back to heating the feed solution.
Distillation processes were the first desalination techniques conducted on a large commercial scale and still account for a large portion of the modern world’s desalination capacity.

The second approach is to physically separate the brine components using an osmotic membrane transparent only for water molecules, which move in response to a water chemical potential difference.
In the context of our study, we will mention the reverse osmosis process (RO)~--- the major process of all the modern desalination industry, and the newly emerging membrane
technology~--- forward osmosis (FO) \cite{Akther2015}. 
In RO, the difference in water chemical potential originates from a difference in pressures applied to the feed and to product solutions. 
In FO, the chemical potential difference is due to an addition to solution from one side of the membrane (draw solution) the so-called draw solutes, which are lowering the water chemical potential in draw solution.

Distillation is easy and cheap technology but it is characterized by relatively high energy costs due to the dissipation of thermal energy. 
On the contrary, RO uses expensive osmotic membranes, which need to be replaced regularly because of scaling and fouling. 
Moreover, RO requires very high operating pressures, ranging from 20 to 200 bar, to let the water pass through the membrane. 
But in terms of energy losses, it works close to the thermodynamic limit.
Thus, theoretically, per one ion pair transferred from freshwater to salty water, it consumes only the energy equal to the difference between the chemical potentials of the transfered ions.

The absence of large hydraulic pressure in FO process (unlike in RO) allows to reduce the energy consumption in pumping, reduce membrane scaling and fouling, and therefore significantly increase the lifetime of the membranes. 
In FO, the draw solutes (agents) are dispersed and/or dissolved in water form homogeneous draw solutions. 
The correct choice of the draw agents is a task of paramount importance. 
As an osmotically driven process, the draw solute is expected to significantly reduce the water chemical potential, and consequently generate a high osmotic pressure. 
On the other hand, the draw solute is expected to be easily separated from water \cite{Cai2016}.

\paragraph{Hydrogels for desalination.}
Hydrogels are three-dimensional networks of polymer chains that are crosslinked by either physical or chemical bonds. 
They are able to entrap large volumes of water attracted by the high concentration of hydrophilic groups. 
When a dehydrated or deswollen hydrogel uptakes water, its polymer chains extend, creating a swelling pressure. 
For example, as reported in \cite{Wack_2009} weakly crosslinked poly(acrylic acid) (PAA) copolymers with polymer volume fractions between 0.03 and 0.30 exhibit a swelling pressure ranging from 0.20--4.23 MPa.
Polyelectrolyte hydrogels, which are carrying ionic groups on the comonomer units (like PAA), are able to reject salt ions from the solution, they absorb a solution of lower salinity than that they are equilibrated with.

An important, advantageous aspect of polymer hydrogels is that they can undergo reversible volume change, \ie solution --- gel phase transitions in response to environmental stimuli. 
This aspect causes hydrogels to be labeled as ‘smart’ hydrogels. 
Many physical and chemical stimuli have been applied to induce various responses of such smart hydrogels, in particular, to change them from hydrophilic to hydrophobic, thereby releasing water. 
The physical stimuli include: temperature, solvent composition, light, mechanical pressure, sound, electric and/or magnetic field, whilst the chemical (or biochemical) stimuli include pH, ionic strength, and specific molecular recognition \cite{Tanaka_1982,Serizawa_2001,Lietor_Santos_2009,Qiu_2001}.

The smart properties of hydrogels were utilized for desalination in a recent study by Li \etal \cite{Li2011}, where hydrogels have been considered as draw agents for FO. 
In this study authors have shown, that hydrogels, 
are able to absorb water across the FO membrane due to their swelling and osmotic pressure and
allow dewatering under the influence of stimuli (thermal and/or mechanical), due to their insoluble cross-linked polymer network.
For instance, Li \etal proposed the use of hydrogel made of thermoresponsive polyelectrolyte~--- copolymer of poly-N-isopropyl acrylamide (p-NIPAAm) and polyacrylic acid (p-AA).
Depending on the temperature the network of the gel appears to be either hydrophilic or hydrophobic. 
In the hydrophilic state, the gel is used as a draw solute accumulating water inside, then in the hydrophobic state, the water releases out.

Moreover, from a thermodynamics perspective, the polyelectrolyte hydrogel itself is an osmotic membrane generating Donnan potential, which rejects ions between outer and inner solutions \cite{Wang_2014}. 
Such a view on hydrogels was employed in a series of works by the group of prof. Wilhelm (see for example \cite{Arens_2017,Fengler_2020}). 
The authors of these works proposed to get rid of the osmotic membrane and simply use only micro filtration to compress the hydrogel squeezing out the accumulated inside the gel solution. 
In their method first, the deswollen hydrogel was equilibrated with saline water feed. 
During equilibration, the gel was swelling absorbing water. 
Then it was taken out from the feed solution and mechanically squeezed by means of a microfiltration membrane. 
The squeezed out brine was obtained to have lower salinity than the feed water.

A similar approach was used by Ali \etal \cite{Ali2015}. 
In their study the thermosensitive gel was used (based on copolymers p-NIPAAm and p-AA), so instead of physical compression, the dewatering was done by an external heating stimulus (sunlight). 
At night the gel was equilibrating with feed water and at daytime, under sunlight, the gel was shrinking and releasing the collected inside solution.

@@

%The maximum entropy principle requires the density of charges to be more-less the same throughout the system medium. 
Since the polyelectrolyte gel has its own charges and its own neutralizing counterions, the density of mobile ions (which are able to freely enter and leave the gel) in gel appears to be smaller than their density outside the gel.
Therefore the internal solution of the gel has a lower density of mobile ions than that in the solution outside. 
In that sense, the gel acts as an osmotic membrane separating solutions \cite{Rud2020}. 
The driving force of the separation is the Donnan potential originated from the charges of the hydrogel network. 
The density of mobile ions difference between ions concentrations in the internal and external solutions
is defined by Donnan law \cite{Rud2018}
\begin{equation*}
\frac{\ccl\gel}{\cs}=\frac{\cs}{\cna\gel}=\sqrt{1+\left(\frac{\alpha}{2\cs\Vgel}\right)^{2}}\pm\frac{\alpha}{2\cs\Vgel}
\end{equation*}
where $\ccl\gel$ and $\cna\gel$ are the internal concentration of monovalent anions and cations, $\cs$ is the salinity of external solution $\Vgel$ is the gel molar volume (inverse density of gel segments) and $\alpha$ is ionization degree. 
The $\pm$ sign in the formula accounts for the change between the cases of gel made of polyanion or polycation.

The lower salinity solution can be extracted from the gel by means of compression and/or other stimulus providing that the charge of the gel remains constant.
In the case of gel made of weak (pH-sensitive) polyelectrolyte, the compression discharges the gel, therefore the neutralizing counterions leave out the gel diminishing the desalination effect \cite{Rud2018}. 
%Nevertheless, in\cite{Rud2018} we discussed how the release of counterions may be employed for desalination as well.

One can argue that the sole Donnan effect is insufficient to achieve a high salt rejection \cite{Cai2016} and the salinity of the water squeezed from hydrogel under a very high hydraulic pressure (up to 100 bar \cite{Fengler_2020}) turns out to be not much different from the initial.
Indeed, the use of high hydraulic pressure diminishes all the advantages of this method over RO, and the reversibility of hydrogel swelling after strong compression remains questionable. 
Nevertheless, in the current study, we model the compression of the gel limiting ourselves to low compression rates, less than 5 bar, and studied how the compression of the gel affects the surrounding salinity.
We demonstrated the desalination process composed of a cascade of following one by each other gel swellings and compressions driving the salinity of gel supernate down to drinkable water. 
%In the current study we would like to concentrate on the use of gel made of strong polyelectrolyte for water desalination. 
%Namely, to investigate how the compression of the gel affects the salinity of the surrounding the gel solution and to propose the way of water desalination employing the hydrogel swelling/deswelling set of processes.

%The target of our study is to model the thermodynamic equilibrium between the two boxes and to show how the compression of the gel affects the ion density in the outside solution.

@@

\begin{figure*}[t]
\subfloat[The gel partial pressure vs gel molar volume \label{fig: PV}]
{\includegraphics[width=0.48\textwidth]{figures/fig_PV_pub}}
\hspace{0.02\textwidth}
\subfloat[Supernate salinity vs gel molar volume\label{fig: CV}]
{\includegraphics[width=0.48\textwidth]{figures/fig_CV_pub}}
\caption{
The compression of the gel in open system (solid lines) and in closed system (dotted lines). 
Each solid curve corresponds to different salinity of the reservoir $\cs$ (see legend). Shadowed area limit the states with applied pressure below zero and above 5 bar.\label{fig: PV and CV}}
\todoi{replace $V$ by $\Vgel$ in Figures}
\end{figure*}

@@

\begin{figure*}[t]
\subfloat[Pressure vs gel molar volume \label{fig: NV}]
{\includegraphics[width=0.49\textwidth]{figures/fig_NV_pub}}
\subfloat[Salinity vs gel molar volume\label{fig: CN}]
{\includegraphics[width=0.495\textwidth]{figures/fig_NVbox_mu_pub}}
\caption{The compression of the gel in open system (solid lines) and in closed
system (dotted lines). Each solid curve corresponds to different salinity
of the reservoir $\cs$ (see legend). Shadowed area limit the states
with applied pressure below zero and above 5 bar.\label{fig: NV and CN}}
%\todoi{Replace y-axis labels by $\ncl$}
\end{figure*}

@@

\caption{ORIGINAL TABLE (to be removed)
}
\end{table}

\end{document}
